---
title: "Introduction to dplyr"
author: "Fabrice Rossi"
format: 
  revealjs:
    echo: false
  pptx:
    echo: false
---

```{r}
here::i_am("dplyr-101.Rproj")
library(here)
```

```{r}
library(ggplot2)
library(dplyr)
```

```{r}
data("diamonds")
```

# Introduction to `filter`

## Link between diamond carat and price

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```

## Selecting a subset of a dataset

```{r}
#| echo: true
diamonds_2ct <- diamonds |>
  filter(abs(carat - 2) <= 0.1)
```

```{r}
ggplot(diamonds_2ct, aes(x = carat, y = price)) +
  geom_point()
```

```{r}
nb_diamonds_near_2 <- diamonds_2ct |> nrow()
nb_diamonds_below_2 <- diamonds_2ct |>
  filter(carat < 2) |> 
  nrow()
```

## Gap in the distribution

There are:

-  `r nb_diamonds_below_2` diamonds with weight between 1.9 and 1.99 carats
-  `r nb_diamonds_near_2-nb_diamonds_below_2` diamonds with weight between 2 and
   2.1 carats

There is a clear gap in the distribution of the weights, with a very small number
of diamonds just before 2 carats compared to the number of them just after 2 carats.


## Prices of 2 carat diamonds

```{r}
diamonds |>
  filter(carat >= 2, carat <= 2.1) |>
  ggplot(aes(x = carat, y = price, group = carat)) +
  geom_boxplot()
```

## Prices of 2 carat diamonds

```{r}
diamonds |>
  filter(carat >= 2, carat <= 2.1) |>
  ggplot(aes(x = carat, y = price, group = carat)) +
  geom_violin()
```
