---
title: "Introduction to dplyr"
author: "Fabrice Rossi"
format: 
  revealjs:
    echo: false
  pptx:
    echo: false
lightbox: true
df-print: kable
---

```{r}
here::i_am("dplyr-101.Rproj")
library(here)
```

```{r}
library(ggplot2)
theme_set(theme_bw())
library(dplyr)
```

```{r}
data("diamonds")
```

# Introduction to `filter`

## Link between diamond carat and price

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```

## Selecting a subset of a dataset

```{r}
#| echo: true
diamonds_2ct <- diamonds |>
  filter(abs(carat - 2) <= 0.1)
```

```{r}
ggplot(diamonds_2ct, aes(x = carat, y = price)) +
  geom_point()
```

```{r}
nb_diamonds_near_2 <- diamonds_2ct |> nrow()
nb_diamonds_below_2 <- diamonds_2ct |>
  filter(carat < 2) |> 
  nrow()
```

## Gap in the distribution

There are:

- `r nb_diamonds_below_2` diamonds with weight between 1.9 and 1.99 carats
- `r nb_diamonds_near_2-nb_diamonds_below_2` diamonds with weight between 2 and 2.1 carats

There is a clear gap in the distribution of the weights, with a very small number of diamonds just before 2 carats compared to the number of them just after 2 carats.

## Prices of 2 carat diamonds

```{r}
diamonds |>
  filter(carat >= 2, carat <= 2.1) |>
  ggplot(aes(x = carat, y = price, group = carat)) +
  geom_boxplot()
```

## Prices of 2 carat diamonds

```{r}
diamonds |>
  filter(carat >= 2, carat <= 2.1) |>
  ggplot(aes(x = carat, y = price, group = carat)) +
  geom_violin()
```

## Using filter {.smaller}

Filter is used as follows

```{r}
#| eval: false
#| echo: true
filter(dataframe, expression with a logical value)
```

Interpretation: the code keeps the rows of the data frame for which the expression evaluates to `TRUE`.

```{r}
filter(diamonds, price == 1000) |>
  ggplot(aes(y=cut)) +
    geom_bar()
```

## Multiple logical expressions {.smaller}

::::: columns
::: {.column width="40%"}
Filter can be used as follows

```{r}
#| eval: false
#| echo: true
filter(dataframe, exp1, exp2, exp3)
```

Interpretation: the code keeps the rows of the data frame for which all expressions evaluate to `TRUE`.
:::

::: {.column width="60%"}
```{r}
filter(diamonds, price == 1000, cut == "Ideal" ) |>
  ggplot(aes(y=color)) +
    geom_bar()
```
:::
:::::

# Logical expressions

## Comparison operators

Everything that compares two values and returns `TRUE` or `FALSE`.

```{r}
#| echo: true
x <- 2
x == 2
x >= 3
```

List: `<`, `<=`, `==`, `>=`, `>` and `!=`

## Logical operators

Combining true value.

```{r}
#| echo: true
TRUE & FALSE
TRUE | FALSE
!FALSE
```

- `|` logical OR
- `&` logical AND
- `!` logical NOT

## Examples

Two equivalent filter operations:
```{r}
#| echo: true
#| eval: false
diamonds |> filter(price == 1000, cut == "Ideal") 
diamonds |> filter( (price == 1000) & (cut == "Ideal") ) 
```

Two different filter operations:
```{r}
#| echo: true
#| eval: false
diamonds |> filter(price == 1000, cut == "Ideal") 
diamonds |> filter( (price == 1000) | (cut == "Ideal") ) 
```
